### 2016-05-22 paper/algorithms.tex
Numerical Exp section expanded, more typos fixed.

\section{Inference Algorithm and Complexity Analysis}\label{sect:alg}
%
Based on the above discussion and the theoretical developments in the last section, we present below a goal-oriented inference algorithm that allows one to combine models of varying fidelity, while maintaining rigorous control of QoI error.
%
\subsection{Goal Oriented Inference Algorithm}
%
Just as error estimates can be used to guide mesh-refinement \cite{BecRann01}, the error estimate (\ref{eq:finErrExp}) can be localized to give elemental contributions and used to guide the division of the domain for a mixed-fidelity model. The error estimate can be calculated again, using the mixed-fidelity model as the lower-fidelity model. This process can be repeated, successively increasing the proportion of the domain in which the high-fidelity model is used, until some threshold is reached. \red{We should include something about how to map error to elements to refine.} We present below an algorithm that accomplishes the above goals.
%
\alglanguage{pseudocode}
\begin{algorithm}[h!]
\small
\caption{An algorithm to adaptively build a mixed-fidelity model for low error in the QoI.}
\label{alg:refSeries}
\begin{algorithmic}[1]
\State{Define maximum acceptable absolute relative QoI error \texttt{errTol}}
\State{Define maximum number of adaptive iterations \texttt{maxIter}}
\Procedure{$\texttt{BuildMF}$}{HF model, LF model, \texttt{errTol}, \texttt{maxIter}}
	\State{Let the model MF$_0$ be the LF model applied everywhere in the domain.}
	\State{$i\gets0$}
	\State{Solve for stationary point $\Psi_{MF_0}$ of augmented Lagrangian $\mathcal{M}_{MF_0}$}
	\State{Solve QoI error adjoint equation, linearized about $\Psi_{MF_0}$, for

@@

Algorithm~\ref{alg:refSeries} is applicable to a large class of models. The lower-fidelity model could, for example, be a simplified model including fewer physical phenomena, be a reduced-order model, or have a reduced parameter space. The two models could also correspond to two levels of mesh-refinement, though in this case the method described in~\cite{BecVex05} could be more efficient, since interpolation could be used to estimate $\Psi_{HF}-\Psi_{LF}$ instead. The derived error estimate is not applicable to all models, however. The two models have to be expressed in a weak form, so this cannot be applied to, for example, a model of chemical reactions using kinetic Monte Carlo. We need some degree of compatibility between the two models; namely, we assume that $\Psi_{LF}$ will be in a space admissible to $\mathcal{M}'_{HF,\Psi}$, and that the QoI functional $I$ is applicable to both $(q_{HF},u_{HF})$ and $(q_{LF},u_{LF})$.
% 
\subsection{Complexity Analysis}
%
We now derive the computational complexity of using algorithm~\ref{alg:refSeries} to solve an inference problem, and compare it to extant strategies. Let the state, adjoint, and parameter variables each have $N$ degrees of freedom, and assume that an 'all at once' strategy is used to solve the KKT system. The cost of single linear solve with such a system is $\orderof{3N}^{\gamma}$, where $\gamma$ depends on the linear solver used~\cite{}. The cost of solving for the auxiliary variables, a linear system, is $\orderof{3N}^{\gamma}$, and the cost for computing the supplementary adjoint is $\orderof{6N}^{\gamma}$.

If one utilizes algorithm~\ref{alg:refSeries} to solve the inference problem, with $T$ adaptive iterations the total computational cost is,
%
\begin{align}
\label{eq:cost_adapt}
C_{MF} &= \sum_{i=1}^{T}L_{i}(3N)^{\gamma} + 1(3N)^{\gamma} + 2^{\gamma}(3N)^{\gamma} \nonumber \\
&= \sum_{i=1}^{T} (L_i + 1 + 2^{\gamma}) (3N)^{\gamma}
\end{align}
% 
where $L_i$ is the number of nonlinear solver steps needed to solve the KKT system for the low or mixed fidelity model. The corresponding cost for a nonadaptive algorithm, for a nonlinear inference problem, solved using a continuation scheme is,
%
\begin{equation}
\label{eq:cost_nadapt}
C_{HF} = \sum_{i=1}^{C}K_{i}(3N)^{\gamma}
\end{equation}
% 
where $K_i$ is the number of nonlinear solver steps needed to solve the KKT system for the high fidelity model, and $C$ is the number of continuation steps.

Comparing Eq.~\eqref{eq:cost_adapt} and~\eqref{eq:cost_nadapt}, we have,
%
\begin{equation}
\label{eq:cost_compare_0}
C_{MF} < C_{HF} \implies \sum_{i=1}^{T} (L_i + 1 + 2^{\gamma}) < \sum_{i=1}^{C}K_{i}
\end{equation}
%
If the low and multi-fidelity models are linear or nonlinear in a very small region $L_i \approx 1$, and we have the following condition,
%
\begin{equation}
\label{eq:cost_compare_1}
T < \frac{\sum_{i=1}^{C}K_{i}}{2 + 2^{\gamma}}
\end{equation}
% 
The bound on $T$ is strictest when $\gamma = 3$, when Gaussian elimination is used, giving $T < \frac{\sum_{i=1}^{C}K_{i}}{10}$.
%
\subsubsection{Illustrative Example}
%
We consider the convection-diffusion-reaction term described in Section \ref{sec:cdvcdr}, with a reaction term $k_r=-616$ in the high-fidelity model. This reaction term is large enough that the Newton solver will not converge with a zero initial guess. We first solve the inverse problem with the low-fidelity model ($k_r=0$), and then use a simple continuation approach, using the solution at one value of $k_r$ as the initial guess for the next.\footnote{not arclength continuation, which is more difficult to implement, though libMesh appears to have a class for such continuation, which we can use if this seems like a viable direction} We increase the reaction term in increments of $\Delta k_r=100$, and halve the increment each time the step is too large (Newton solver does not converge at the next $k_r$ value). From $k_r=0$ to $k_r=-616$, this results in 9 continuation steps being taken, with $\sum_{i=1}^{C} K_i=30$. With Gaussian elimination chosen as the linear solver, this gives us $T < 3$, i.e.\ a budget of 2 adaptive steps. Using algorithm~\ref{alg:refSeries} with 2 refinement steps, and only $10\%$ of the domain refined to use the high-fidelity model, the estimated relative error is $<1\%$. Indeed, the ratio $\frac{C_{MF}}{C_{HF}}$ was $\frac{24}{30}$, indicating about $20\%$ reduction in computational cost, with the worst case linear solver used.

In other words, for a highly nonlinear problem, using algorithm~\ref{alg:refSeries}, even with the worst case linear solver, one can get to $<1\%$ error in the QoI, while avoiding any need for continuation to handle nonlinearities.

>>>>

\section{Inference Algorithm and Complexity Analysis}\label{sect:alg}
%
Based on the theoretical developments in the last section, we present below a goal-oriented inference algorithm that allows one to combine models of varying fidelity, while maintaining rigorous control of QoI error.
%
\subsection{Goal Oriented Inference Algorithm}
%
Just as error estimates can be used to guide mesh-refinement \cite{BecRann01}, the error estimate (\ref{eq:finErrExp}) can be localized to give elemental contributions and used to guide the division of the domain for a mixed-fidelity model. The error estimate can be calculated again, using the mixed-fidelity model as the lower-fidelity model. This process can be repeated, successively increasing the proportion of the domain in which the high-fidelity model is used, until some threshold is reached. \red{We should include something about how to map error to elements to refine.} We present below an algorithm that accomplishes the above goals.
%
\alglanguage{pseudocode}
\begin{algorithm}[h!]
\small
\caption{An algorithm to adaptively build a mixed-fidelity model for low error in the QoI.}
\label{alg:refSeries}
\begin{algorithmic}[1]
\State{Define maximum acceptable absolute relative QoI error \texttt{errTol}}
\State{Define maximum number of adaptive iterations \texttt{maxIter}}
\Procedure{$\texttt{BuildMF}$}{HF model, LF model, \texttt{errTol}, \texttt{maxIter}}
	\State{Let the model MF$_0$ be the LF model applied everywhere in the domain.}
	\State{$i\gets0$}
	\State{Solve for stationary point $\Psi_{MF_0}$ of augmented Lagrangian $\mathcal{M}_{MF_0}$}
	\State{Solve QoI error adjoint equation, linearized about $\Psi_{MF_0}$, for

@@

Algorithm~\ref{alg:refSeries} is applicable to a large class of models. The lower-fidelity model could, for example, be a simplified model including fewer physical phenomena, be a reduced-order model, or have a reduced parameter space. The two models could also correspond to two levels of mesh-refinement, though in this case the method described in~\cite{BecVex05} could be more efficient, since interpolation could be used to estimate $\Psi_{HF}-\Psi_{LF}$ instead. The derived error estimate is not applicable to all models, however. The two models have to be expressed in a weak form, so this cannot be applied to, for example, a model of chemical reactions using kinetic Monte Carlo. We need some degree of compatibility between the two models; namely, we assume that $\Psi_{LF}$ will be in a space admissible to $\mathcal{M}'_{HF,\Psi}$, and that the QoI functional $I$ is applicable to both $(q_{HF},u_{HF})$ and $(q_{LF},u_{LF})$.
% 
\subsection{Complexity Analysis}\label{sect:alg_complexity}
%
We now derive the computational complexity of using algorithm~\ref{alg:refSeries} to solve an inference problem, and compare it to extant strategies. Let the state, adjoint, and parameter variables each have $N$ degrees of freedom, and assume that an `all at once' strategy is used to solve the KKT system. The cost of single linear solve with such a system is $\orderof{3N}^{\gamma}$, where $\gamma$ depends on the linear solver used~\cite{}. The cost of solving for the auxiliary variables, a linear system, is $\orderof{3N}^{\gamma}$, and the cost for computing the supplementary adjoint is $\orderof{6N}^{\gamma}$.

If one utilizes algorithm~\ref{alg:refSeries} to solve the inference problem, with $T$ adaptive iterations the total computational cost is,
%
\begin{align}
\label{eq:cost_adapt}
C_{MF} &= \sum_{i=1}^{T}L_{i}(3N)^{\gamma} + 1(3N)^{\gamma} + 2^{\gamma}(3N)^{\gamma} \nonumber \\
&= \sum_{i=1}^{T} (L_i + 1 + 2^{\gamma}) (3N)^{\gamma}
\end{align}
% 
where $L_i$ is the number of nonlinear solver steps needed to solve the KKT system for the low or mixed fidelity model at each adaptive step. The corresponding cost for solving the high fidelity inference problem, with a nonadaptive algorithm, based on a continuation scheme is,
%
\begin{equation}
\label{eq:cost_nadapt}
C_{HF} = \sum_{i=1}^{C}K_{i}(3N)^{\gamma}
\end{equation}
% 
where $K_i$ is the number of nonlinear solver steps needed to solve the KKT system for the high fidelity model at each continuation step, and $C$ is the number of continuation steps.

Comparing Eq.~\eqref{eq:cost_adapt} and~\eqref{eq:cost_nadapt}, we have,
%
\begin{equation}
\label{eq:cost_compare_0}
C_{MF} < C_{HF} \implies \sum_{i=1}^{T} (L_i + 1 + 2^{\gamma}) < \sum\limits_{i=1}^{C}K_{i}
\end{equation}
%
If the low and multi-fidelity models are linear or nonlinear in a very small region $L_i \approx 1$, and we have the following condition,
%
\begin{equation}
\label{eq:cost_compare_1}
T < \frac{\sum\limits_{i=1}^{C}K_{i}}{2 + 2^{\gamma}}
\end{equation}
% 
The bound on $T$ will be strictest if Gaussian elimination is used, with $\gamma = 3$, giving $T < \frac{\sum\limits_{i=1}^{C}K_{i}}{10}$.
%

### 2016-05-22 paper/mathematical_formulation.tex
Numerical Exp section expanded, more typos fixed.

In the case of a goal-oriented inverse problem, the ultimate purpose of inferring the unknown parameters is to calculate some Quantity of Interest (QoI). Assuming a single scalar QoI, we denote this QoI by $I(q,u)$, where $I:Q\times U\to\R$ is a functional that maps the parameters and state to our QoI. Since our ultimate goal is to compute this QoI, the particular tradeoff we consider is the fidelity and resultant computational expense of the model we use, versus the error in the QoI. We thus seek to introduce the QoI functional $I$ into the inverse problem formulation, by introducing auxilary variables and additional adjoint equations. Then we use this formulation to derive an a posteriori error estimate for $I$, where the errors considered are those which occur due to the use of constraints other than those given by Eq.~\eqref{eq:invOpt_cons}.
%
%------------------------------------------------------------------------------%
\subsection[Error Estimate for a Goal-Oriented Inverse Problem]{Error Estimate for a Goal-Oriented Inverse \\Problem}  \label{sec:deriv}
%------------------------------------------------------------------------------%
%
For a given hierarchy of models, consider the QoI calculated from inferring the parameters with the highest-fidelity model; we take this QoI to be the value with which we compare other QoI estimates. In this section we derive an a posteriori estimate for the error in the QoI from inferring the parameters with a lower-fidelity model, as compared to that which would have resulted from solving the inverse problem with the highest-fidelity model. The main result is,
%
\begin{theorem}
\label{thm:error_estimate}
Following the notation established in section~\ref{sec:setup}, let the bilinear form $a:U \times U \to \Reals$ be three times continuously differentiable with respect to the state $u$ and parameters $q$. Let the observation operator $C:U\to\Reals^{n_d}$ be three times continuously differentiable with respect to the state $u$. Also, let the regularization operator $R:Q\to\Reals$ be differentiable with respect to the parameter $q$, and the functional $I:Q\times U\to\Reals$ be differentiable with respect to the state $u$ and parameter $q$.

@@

Denote by $\Psi_{HF}$ and $\Psi_{LF}$, stationary points of the high and low fidelity versions of Eq.~\eqref{eq:InvsOpt_auglag} and, consider the adjoint problem,
%
\begin{equation}
\label{eq:superAdjEq}
\mathcal{M}'_{HF,\Psi}(\Lambda;\Psi_{HF})(\Phi)=\mathcal{Q}(\Phi)=\mathcal M'_{HF}(\Psi_{LF})(\Phi),\quad\forall\Phi\in(Q_{HF}\times U_{HF}\times U_{HF})^2,
\end{equation}
%
for the supplementary adjoint $\Lambda$. Then, the error in the Quantity of Interest $I$ is given by,
%
\begin{multline}
\label{eq:finErrExp}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\-\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Lambda)+\mathcal M_{HF}(\Psi_{LF})-\mathcal M_{LF}(\Psi_{LF})+\mathcal{R}(e^3).
\end{multline}
%
\end{theorem}
%
\begin{proof}
%
Observe that,
%
\begin{equation}
\label{eq:MeqI}
\mathcal{M}(\Psi)=I(q,u),
\end{equation}
%
since taking variations of $\mathcal{M}$ with respect to the auxiliary variables gives that $\xi_\Psi$ is a stationary point of $\mathcal{L}$.

Extending the property in Equation (\ref{eq:MeqI}) to the augmented Lagrangians for the high and low fidelity models, we have,
%
\begin{multline}
\label{eq:repIwithM}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF})+\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})\textrm{.}
\end{multline}
%
Applying Proposition 3 from~\cite{BecVex05} for the difference $\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF})$,
\begin{equation}
\label{eq:beckvex}
\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF}) = \frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})+\mathcal{R}(e^3)\textrm{,}
\end{equation}
where $\mathcal{R}$ is a remainder term that is third-order in the error $e=\Psi_{HF}-\Psi_{LF}$. Combining Equations (\ref{eq:repIwithM}) and (\ref{eq:beckvex}) we obtain
\begin{multline}
\label{eq:preadj}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})+\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})+\mathcal{R}(e^3)\textrm{.}
\end{multline}

>>>>

In the case of a goal-oriented inverse problem, the ultimate purpose of inferring the unknown parameters is to calculate some Quantity of Interest (QoI). Assuming a single scalar QoI, we denote this QoI by $I(q,u)$, where $I:Q\times U\to\R$ is a functional that maps the parameters and state to our QoI. Since our ultimate goal is to compute this QoI, the particular tradeoff we consider is the fidelity and resultant computational expense of the model we use, versus the error in the QoI. We thus seek to introduce the QoI functional $I$ into the inverse problem formulation, by introducing auxilary variables and additional adjoint equations. Then we use this formulation to derive an a posteriori error estimate for $I$, where the errors considered are those which occur due to the use of constraints other than those given by Eq.~\eqref{eq:invOpt_cons}.
%
%------------------------------------------------------------------------------%
\subsection[Error Estimate for a Goal-Oriented Inverse Problem]{Error Estimate for a Goal-Oriented Inverse \\Problem}  \label{sec:deriv}
%------------------------------------------------------------------------------%
%
For a given hierarchy of models, consider the QoI calculated from inferring the parameters with the highest-fidelity model; we take this QoI to be the value with which we compare other QoI estimates. In this section we derive an a posteriori estimate for the error in the QoI from inferring the parameters with a lower-fidelity model. The main result is,
%
\begin{theorem}
\label{thm:error_estimate}
Following the notation established in section~\ref{sec:setup}, let the bilinear form $a:U \times U \to \Reals$ be three times continuously differentiable with respect to the state $u$ and parameters $q$. Let the observation operator $C:U\to\Reals^{n_d}$ be three times continuously differentiable with respect to the state $u$. Also, let the regularization operator $R:Q\to\Reals$ be differentiable with respect to the parameter $q$, and the functional $I:Q\times U\to\Reals$ be differentiable with respect to the state $u$ and parameter $q$.

@@

Denote by $\Psi_{HF}$ and $\Psi_{LF}$, stationary points of the high and low fidelity versions of Eq.~\eqref{eq:InvsOpt_auglag} and, consider the adjoint problem,
%
\begin{equation}
\label{eq:superAdjEq}
\mathcal{M}'_{HF,\Psi}(\Lambda;\Psi_{HF})(\Phi)=\mathcal{Q}(\Phi)=\mathcal M'_{HF}(\Psi_{LF})(\Phi),\quad\forall\Phi\in(Q_{HF}\times U_{HF}\times U_{HF})^2,
\end{equation}
%
for the supplementary adjoint $\Lambda$. Then, the error in the Quantity of Interest $I$ is given by,
%
\begin{multline}
\label{eq:finErrExp}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\-\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Lambda)+\mathcal M_{HF}(\Psi_{LF})-\mathcal M_{LF}(\Psi_{LF})+\mathcal{R}(e^3).
\end{multline}
%
where $\mathcal{R}$ is a remainder term that is third-order in the error $e=\Psi_{HF}-\Psi_{LF}$.
\end{theorem}
%
\begin{proof}
%
Observe that,
%
\begin{equation}
\label{eq:MeqI}
\mathcal{M}(\Psi)=I(q,u),
\end{equation}
%
since taking variations of $\mathcal{M}$ with respect to the auxiliary variables gives that $\xi_\Psi$ is a stationary point of $\mathcal{L}$.

Extending the property in Equation (\ref{eq:MeqI}) to the augmented Lagrangians for the high and low fidelity models, we have,
%
\begin{multline}
\label{eq:repIwithM}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF})+\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})\textrm{.}
\end{multline}
%
Applying Proposition 3 from~\cite{BecVex05} for the difference $\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF})$,
\begin{equation}
\label{eq:beckvex}
\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF}) = \frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})+\mathcal{R}(e^3)\textrm{,}
\end{equation}
Combining Equations (\ref{eq:repIwithM}) and (\ref{eq:beckvex}) we obtain
\begin{multline}
\label{eq:preadj}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})+\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})+\mathcal{R}(e^3)\textrm{.}
\end{multline}
